An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-23 paper.tex
Clean up presentation of algebraic BFS

@@

\section{Formulating the Evolving Graph BFS with Linear Algebra}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ time stamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) =
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$.
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle
\]

This algebraic formulation exploits a graphical interpretation of matrix--vector
products involving the adjacency matrix~\cite{kegi11}.
Let $G = (V, E)$ be a (directed or undirected) graph
and $k$ be a node in the graph $G$. If $A$ is the adjacency matrix of $G$ and
$b$ is the column vector
\[
b_i =
\begin{cases}
1 & \mbox{if $i = k$,} \\
0 & \mbox{otherwise,}
\end{cases}
\]
then $A^Tb$ is a column vector whose nonzero entries have indices that are the
forward-neighbors of $k$.

The evolving graph BFS can be thought of as a generalization of blocked
matrix--vector products.
Introduce a new operator $\odot$, a variation on the ordinary matrix--vector
product such that
\[
A^T \odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
Then, the forward-neighbors of node $(k, 1)$ in $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ are the nonzero indices of the
elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Tb, \quad  (A^{[2]})^T\odot b, \quad \ldots \quad (A^{[n]})^T\odot b \big\rangle.
\end{equation}
The forward-neighbors at time stamp $t$ are the nonzero indices of the $t$-th element of the above sequence.
Hence given $A_n$ and a vector $b$, we can use \eqref{eq:out_ne} to
get a $1$-step BFS tree. In Figure \ref{fig:eg_shortest_path}, the forward-neighbors
of node $(1, t_1)$ can be computed by

\begin{align*}
\left\langle
  \begin{bmatrix}
    0 & 0 & 0 \\
    1 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix}\!\!\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
   0 & 0 & 0 \\
   0 & 0 & 0 \\
   1 & 0 & 0
 \end{bmatrix}\!\!
\odot\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 0 \\
  0 & 1 & 0
 \end{bmatrix}\!\!
\odot\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
\right\rangle
\\
=
\left\langle
\begin{bmatrix}
0 \\
1 \\
0
\end{bmatrix},~
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix},~
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
\right\rangle
\end{align*}
This means node $(2,t_1)$ and $(1,t_2)$ are the forward-neighbors of $(1,t_1)$.

>>>>

The preceding example illustrates the fact that $G^{[1]}$ plays no part in the
BFS traversal of $G_n$ starting from $(1, t_2)$. In general, all $G^{[t]}$ with
time stamps $t < t'$ for a starting node $(v, t')$ are irrelevant to the BFS
traversal. Hence without loss of generality we may assume that BFS is always
computed with a root at time $t_1$, the earliest time stamp in $G_n$.

@@

\section{Formulating the Evolving Graph BFS with Linear Algebra}
\label{sec:representation}

For each static graph $G^{[t]} = (V^{[t]}, E^{[t]})$ that constitutes the
evolving graph $G_n$, define its corresponding
$\left\vert V^{[t]}\right\vert \times \left\vert V^{[t]}\right\vert$
adjacency matrix with elements
\[
A^{[t]}_{ij} =
\begin{cases}
 1 & \mbox{if $(i,j)\in E^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]

We can then represent $G_n$ using the sequence of adjacency matrices
$A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$. The example in
Figure~\ref{fig:eg_shortest_path} can be represented as
\[
\left\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\right\rangle.
\]

This algebraic representation allows us to exploit a graphical interpretation of
matrix--vector products involving the adjacency matrix~\cite{kegi11}.
If $A$ is the adjacency matrix of a (static) graph $G$ and $e_k$ is $k$th
elementary unit vector, then the nonzero entries of $A^T e_k$ have indices that
are neighbors of $k$.
The algebraic formulation of BFS on evolving graphs follows similarly, but
requires a new kind of matrix--vector product, $\odot$, defined by
\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ are then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas all other nonzero entries represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

Referring back to the example of Figure~\ref{fig:eg_shortest_path}, the forward
neighbors of node $(1, t_1)$ can be computed by
\begin{align*}
\left\langle
  \begin{bmatrix}
    0 & 0 & 0 \\
    1 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix}\!\!\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}\!,
 \begin{bmatrix}
   0 & 0 & 0 \\
   0 & 0 & 0 \\
   1 & 0 & 0
 \end{bmatrix}\!\!
\odot\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}\!,
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 0 \\
  0 & 1 & 0
 \end{bmatrix}\!\!
\odot\!\!
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}\!
\right\rangle
\\
=
\left\langle
\begin{bmatrix}
0 \\
1 \\
0
\end{bmatrix},~
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix},~
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
\right\rangle
\end{align*}
From this computation, we can deduce that $(2,t_1)$ and $(1,t_2)$ are the forward
neighbors of $(1,t_1)$.
